08_relative_N_inputs

### 2025-07-08 01_tidying_data.qmd
Tidying scripts
All changes

---
title: "Tidying data"
format: html
editor: source
---

@@

aquamaps_0.5d <- data.table::fread(file.path(rdsi_data_dir, "aquamaps/aquamaps_0.6_depth_prepped.csv"))

>>>>

---
title: "Tidying data"
author: Richard Cottrell
date: today
editor: source
---

@@

```{r}
aquamaps_0.5d <- data.table::fread(file.path(rdsi_data_dir, "aquamaps/aquamaps_0.6_depth_prepped.csv"))

### 2025-07-08 02_extracting_production_locations.qmd
Tidying scripts
All changes

---
title: "Extracting production locations for each species to obtain environmental forcings"
format: html
editor: source
---

@@

```{r setup}
library(tidyr)
library(dplyr)
library(magrittr)
library(here)
library(janitor)
library(sf)
library(qs)
library(countrycode)
library(terra)

source(here("00_dir.R"))

species <- c("atlantic_salmon", 
             "general_salmonid", 
             "european_seabass", 
             "gilthead_seabream", 
             "large_yellow_croaker", 
             "japanese_seabass",
             "japanese_amberjack",
             "general_grouper",
             "pompano",
             "milkfish")
```

Isolate production data for species of interest

```{r}
aquaculture_raw <- read_csv(file.path(rdsi_dir, "raw_data/fao/FAO_fishstat_2020/global_aquaculture_production_1950_2020.csv"))

>>>>

---
title: "Extracting production locations for each species to obtain environmental forcings"
author: Richard Cottrell
date: today
editor: source
---

@@

```{r setup}
library(tidyr)
library(dplyr)
library(magrittr)
library(here)
library(janitor)
library(sf)
library(qs)
library(countrycode)
library(terra)
library(units)

source(here("00_dirs.R"))
```

Isolate production data for Atlantic salmon.
```{r rich-raw-processing}
#| eval: false

aquaculture_raw <- read_csv(file.path(rdsi_dir, "raw_data/fao/FAO_fishstat_2020/global_aquaculture_production_1950_2020.csv"))

### 2025-07-08 04_extracting_temperatures.qmd
All changes

The purpose of this markdown is to:

>>>>

The purpose of this markdown is to extract daily temperature timeseries from SST data to drive fish growth in the correct places. The steps in this markdown are:

1. Pull farm locations data (previously determined from FAO stocking locations)
2. Record whether the farm is in the southern or north hemisphere (determines cohort stocking date)
3. Pull SST data from .tif file and construct "typical" temperature year for each farm location
4. Gapfill missing temperature data using closest farm in the same FAO stocking area
5. Save temperature data both with and without geometry information
6. Determine which farms have a mean temperature <=6 C (will be omitted)

### 2025-07-08 08_relative_N_inputs.qmd
All changes

source("00_dirs.R")

overwrite <- F
```

@@

Data for nutrient runoff from agriculture fertilisers and pesticides was downloaded from @halpern_recent_2019-1 [dataset here: @halpern_recent_2019].

@@

halpern_insets <- molle_insets %>% 
  lapply(function(ins) {project(halpern_N_rast, ins, method = "cubic")})

# Change to the Molleweide projection
tuholske_N_rast <- file.path(input_Ndata_path, "tuholske_2021", "global_effluent_2015_tot_N.tif") %>% rast() 
tuholske_insets <- molle_insets %>% 
  lapply(function(ins) {project(tuholske_N_rast, ins, method = "cubic")})

# The halpern data has NA values for land but the tuholske data does not
tuholske_insets <- map2(tuholske_insets, halpern_insets, function(rast, mask) {
  mask(rast, mask)
})

background_N_insets <- map2(halpern_insets, tuholske_insets, function(dat1, dat2) {dat1 + dat2})

map2(names(halpern_insets), background_N_insets, function(nm, rast) {
  writeRaster(rast, file.path(input_Ndata_path, str_c("background_N_inset_", nm, ".tif")))
})
```

@@

# Change to the Molleweide projection
tuholske_N_rast <- file.path(input_Ndata_path, "tuholske_2021", "global_effluent_2015_tot_N.tif") %>% rast() 
tuholske_insets <- molle_insets %>% 
  lapply(function(ins) {project(tuholske_N_rast, ins, method = "cubic")})

# The halpern data has NA values for land but the tuholske data does not
tuholske_insets <- map2(tuholske_insets, halpern_insets, function(rast, mask) {
  mask(rast, mask)
})
```

```{r save-back-N}
#| code-summary: Combine and save background N data

@@

map2(names(halpern_insets), background_N_insets, function(nm, rast) {
  writeRaster(rast, file.path(input_Ndata_path, str_c("background_N_inset_", nm, ".tif")))
})
```

@@

# Load modelled N data

@@

# Combine with background N

```{r}
# 2. Extract 3x3 neighborhood data for each farm point
# Get cell numbers for each farm point
farm_cells <- cellFromXY(back_N_rast, st_coordinates(farm_sfs))

# Function to get 5x5 neighborhood cell numbers
get_neighborhood_cells <- function(raster, center_cells) {
  # Get raster dimensions
  nrows <- nrow(raster)
  ncols <- ncol(raster)

  # Convert cell numbers to row/col indices
  rows <- rowFromCell(raster, center_cells)
  cols <- colFromCell(raster, center_cells)

  # Create all 25 neighbor offsets (including center) for 5x5 grid
  offsets <- expand.grid(row_offset = -2:2, col_offset = -2:2)

  # Initialize list to store results
  all_neighbors <- vector("list", length(center_cells))

  for(i in seq_along(center_cells)) {
    # Calculate neighbor row/col positions
    neighbor_rows <- rows[i] + offsets$row_offset
    neighbor_cols <- cols[i] + offsets$col_offset

    # Check bounds and filter valid cells
    valid_cells <- neighbor_rows >= 1 & neighbor_rows <= nrows & 
                   neighbor_cols >= 1 & neighbor_cols <= ncols

    # Convert back to cell numbers
    neighbor_cells <- cellFromRowCol(raster, 
                                   neighbor_rows[valid_cells], 
                                   neighbor_cols[valid_cells])

    # Create position labels for 5x5 grid
    row_names <- c("T2", "T1", "Crow", "B1", "B2")
    col_names <- c("L2", "L1", "Ccol", "R1", "R2")

    position_grid <- outer(row_names, col_names, paste, sep = "_")
    position_labels <- as.vector(position_grid)
    valid_positions <- position_labels[valid_cells]

    all_neighbors[[i]] <- data.frame(
      farm_id = farm_sfs$farm_id[i],
      cell_id = neighbor_cells,  # for overlap detection
      position = valid_positions,
      stringsAsFactors = FALSE
    )
  }
  return(do.call(rbind, all_neighbors))
}

# Get all neighborhood cells
neighborhood_data <- get_neighborhood_cells(back_N_rast, farm_cells) %>% 
  mutate(
    category = case_when(position == "Crow_Ccol" ~ "farm_direct",
                         position %in% c("T1_L1", "T1_Ccol", "T1_R1", "B1_L1", "B1_Ccol", "B1_R1", "Crow_L1", "Crow_R1") ~ "adjacent_1",
                         T ~ "adjacent_2"),
    category = factor(category, levels = c("farm_direct", "adjacent_1", "adjacent_2")),
    position = factor(position, levels = c(
      "T2_L2",   "T2_L1",   "T2_Ccol",   "T2_R1",   "T2_R2", 
      "T1_L2",   "T1_L1",   "T1_Ccol",   "T1_R1",   "T1_R2", 
      "Crow_L2", "Crow_L1", "Crow_Ccol", "Crow_R1", "Crow_R2", 
      "B1_L2",   "B1_L1",   "B1_Ccol",   "B1_R1",   "B1_R2", 
      "B2_L2",   "B2_L1",   "B2_Ccol",   "B2_R1",   "B2_R2"
    )))

# Extract nutrient values and combine with cell_ids
nutrient_values <- terra::extract(back_N_rast$nutrient_pollution_2013_raw, neighborhood_data$cell_id)
neighborhood_data <- cbind(neighborhood_data, nutrient_values) %>% 
  rename(np = nutrient_pollution_2013_raw)

# Fill gaps from original data with 8 neighbours
neighborhood_data_gaps <- neighborhood_data %>% filter(is.na(np))
for (i in 1:nrow(neighborhood_data_gaps)) {
  adj_cells <- adjacent(back_N_rast$nutrient_pollution_2013_raw, cells = neighborhood_data_gaps$cell_id[i], directions = 8)
  neighborhood_data_gaps$np[i] <- back_N_rast$nutrient_pollution_2013_raw[adj_cells[1,]] %>% 
    pull(nutrient_pollution_2013_raw) %>% 
    mean(na.rm = T)
}
neighborhood_data <- neighborhood_data %>% 
  filter(!is.na(np)) %>% 
  rbind(neighborhood_data_gaps)
neighborhood_data %>% filter(is.na(np)) %>% nrow() # 3221 still blank
neighborhood_data %>% filter(is.na(np) & position == "Crow_Ccol") %>% nrow() # 87 of those are direct farm cells

# Fill gaps from original data with 16 neighbours
neighborhood_data_gaps <- neighborhood_data %>% filter(is.na(np))
for (i in 1:nrow(neighborhood_data_gaps)) {
  adj_cells <- adjacent(back_N_rast$nutrient_pollution_2013_raw, cells = neighborhood_data_gaps$cell_id[i], directions = 16)
  neighborhood_data_gaps$np[i] <- back_N_rast$nutrient_pollution_2013_raw[adj_cells[1,]] %>% 
    pull(nutrient_pollution_2013_raw) %>% 
    mean(na.rm = T)
}
neighborhood_data <- neighborhood_data %>% 
  filter(!is.na(np)) %>% 
  rbind(neighborhood_data_gaps)
neighborhood_data %>% filter(is.na(np)) %>% nrow() # 4489 still blank
neighborhood_data %>% filter(is.na(np) & position == "Crow_Ccol") %>% nrow() # 149 of those are direct farm cells
```

```{r summary}
# Summary statistics
summary_stats <- neighborhood_data %>%
  group_by(farm_id, category) %>%
  reframe(
    mean = mean(np, na.rm = T),
    min = min(np, na.rm = T),
    max = max(np, na.rm = T),
    cells = n()
  )

# Maybe use adjacent farms to fill gaps?

ggplot(neighborhood_data, aes(x = category, y = log(np), fill = position)) +
  geom_boxplot()
```

# Plot mean nutrient pollution

```{r}
farm_N <- summary_stats %>% 
  filter(category == "farm_direct")

farm_N_sfs <- st_sf(farm_N = farm_N$mean, geometry = farm_locs$geometry) %>% 
  st_transform(crs = crs(back_N_rast))

plot(farm_N_sfs)
```

>>>>

source("00_dirs.R")
feed_pal <- c("past" = "#E41A1C", "reference" = "#377EB8", "future" = "#4DAF4A")

prettyplot <- function() {
  theme_classic() +
    theme(legend.position = "none",
          text = element_text(family = "serif", size = 12, colour = "black"),
          axis.title.x = element_text(vjust = 0.5),
          axis.title.y = element_text(hjust = 0.5))
}
```

@@

Data for nutrient runoff from agriculture fertilisers and pesticides was downloaded from @halpern_recent_2019-1 [dataset here: @halpern_recent_2019]. The data is in annual tonnes N.

@@

halpern_insets <- molle_insets %>% 
  lapply(function(ins) {project(halpern_N_rast, ins, method = "cubic")})
```

@@

tuholske_N_rast <- file.path(input_Ndata_path, "tuholske_2021", "global_effluent_2015_tot_N.tif") %>%
  rast() 
tuholske_N_rast <- tuholske_N_rast * 10^-6 # Change from g to t

# Change to the Molleweide projection with insets
tuholske_insets <- molle_insets %>% 
  lapply(function(ins) {project(tuholske_N_rast, ins, method = "cubic")})

# The halpern data has NA values for land but the tuholske data does not, so mask
tuholske_insets <- map2(tuholske_insets, halpern_insets, function(rast, mask) {
  mask(rast, mask)
})
```

```{r save-back-N}
#| code-summary: Combine and save background N data
#| eval: false

@@

map2(names(halpern_insets), background_N_insets, function(nm, rast) {
  writeRaster(rast, 
              file.path(input_Ndata_path, str_c("background_N_inset_", nm, ".tif")), 
              overwrite = T)
})
```

The code above was only run once, background N rasters can now be loaded.

@@

# Combine modelled N data with background N

@@

```{r matching cells}
#| code-summary: Get background nitrogen data for farm cells

farm_cells <- map(background_N_insets, function(back_N_rast) {
  # Get cell numbers for each farm point
  farm_cells <- cellFromXY(back_N_rast, st_coordinates(farm_sfs))
  farm_cells <- farm_locs %>% 
    mutate(cell = farm_cells) %>% 
    filter(!is.na(cell))

  # Extract nutrient values and combine with cell_ids
  farm_cells$back_N <- terra::extract(back_N_rast$nutrient_pollution_2013_raw,
                                      farm_cells$cell)$nutrient_pollution_2013_raw

  # Fill gaps from original data with 8 neighbours
  farm_cells_gaps <- farm_cells %>% filter(is.na(back_N))
  for (i in 1:nrow(farm_cells_gaps)) {
    adj_cells <- adjacent(back_N_rast$nutrient_pollution_2013_raw, 
                          cells = farm_cells_gaps$cell[i], directions = 8)
    adj_cells_N <- back_N_rast$nutrient_pollution_2013_raw[adj_cells[1,]] %>% 
      pull(nutrient_pollution_2013_raw) %>% 
      mean(na.rm = T)
    farm_cells_gaps$back_N[i] <- adj_cells_N
  }
  farm_cells <- farm_cells %>% 
    filter(!is.na(back_N)) %>% 
    rbind(farm_cells_gaps)

  # Fill gaps from original data with 16 neighbours
  farm_cells_gaps <- farm_cells %>% filter(is.na(back_N))
  if (nrow(farm_cells_gaps) != 0) {
      for (i in 1:nrow(farm_cells_gaps)) {
      adj_cells <- adjacent(back_N_rast$nutrient_pollution_2013_raw, 
                            cells = farm_cells_gaps$cell[i], directions = 16)
      adj_cells_N <- back_N_rast$nutrient_pollution_2013_raw[adj_cells[1,]] %>% 
        pull(nutrient_pollution_2013_raw) %>% 
        mean(na.rm = T)
      farm_cells_gaps$back_N[i] <- adj_cells_N
    }
    farm_cells <- farm_cells %>% 
      filter(!is.na(back_N)) %>% 
      rbind(farm_cells_gaps)
  }

  default_val <- exp(median(log(farm_cells$back_N), na.rm = T))
  farm_cells$back_N[is.na(farm_cells$back_N)] <- default_val
  farm_cells$back_N[farm_cells$back_N < 0] <- default_val
  farm_cells
}) %>% 
  bind_rows()
```

```{r combine with N input from farms}

N_input <- output_cohorts_data_path %>% list.files(full.names = T) %>% 
  str_subset("allfarms") %>% 
  str_subset("P_excr") %>% 
  qs::qread() %>% 
  mutate(mean = mean * 6.25) %>% 
  group_by(farm_ID, feed) %>% 
  reframe(mean = sum(mean) * 10^-6) %>% 
  pivot_wider(names_from = feed, values_from = mean)

N_comp <- merge(farm_cells, N_input, by.y = "farm_ID", by.x = "farm_id", all = T) %>% 
  select(-cell) %>% 
  mutate(reference = reference/back_N,
         past = past/back_N,
         future = future/back_N) %>% 
  pivot_longer(names_to = "feed", names_transform = list(feed = as.factor),
               values_to = "ratio", cols = c("reference", "past", "future")) %>% 
  mutate(feed = factor(feed, levels = c("reference", "past", "future")))
```

# Analyse

```{r summary-plots}

N_comp %>%
  group_by(country, farm_id, feed) %>% 
  reframe(mean = mean(ratio, na.rm = T)) %>% 
  pivot_wider(names_from = feed, values_from = mean) %>% 
  mutate(past = (past-reference)/reference,
         future = (future-reference)/reference) %>% 
  select(-reference) %>% 
  pivot_longer(names_to = "feed", values_to = "mean_ratio", cols = c(past, future)) %>% 
  ggplot(aes(x = country, y = mean_ratio, fill = feed)) +
  geom_boxplot() +
  prettyplot() +
  scale_y_continuous(breaks = seq(0,1,0.2), labels = seq(0,100,20), limits = c(0,0.6)) +
  scale_fill_manual(values = feed_pal) +
  labs(y = "Mean change in N input ratio", x = "Country")

N_comp %>% 
  filter(log(ratio) < 20) %>%
  ggplot(aes(x = country, y = log(ratio), fill = feed)) +
  geom_boxplot() +
  prettyplot() +
  # scale_y_continuous(breaks = seq(0,1,0.2), labels = seq(0,100,20), limits = c(0,0.6)) +
  scale_fill_manual(values = feed_pal) +
  labs(y = "Mean change in N input ratio", x = "Country")
```
